### 2018-01-25 submission/manuscript.Rmd
ada background edits

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying biomarkers or individual microbes but did so on a small total number of individuals and only investigated stool.

Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the candidate genera seem to be numerous. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, *Escherichia/Shigella* and *Streptococcus* also have good *in vivo* evidence that individual species within the genus can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera. Some groups have focused on more broad scale community metrics and have found that metrics such as richness are decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching this sample size alone, it is hard to know how extrapolatable these findings truly are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further, since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].

Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to the community on the adenoma or carcinoma. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the tissue directly may provide clearer answers but is not without limitations. Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these many differences that could arise between stool and tissue, an additional question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor or if select members can recapitulate the same information for both adenoma and carcinoma. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the top 5 increased and top 5 decreased RR genera combined resulted in better model classification, based on the area under the curve (AUC). Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification. Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying microbial signatures of CRC (biomarkers) but did so on a small total number of individuals and only investigated stool.

Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and CRC, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. There are also genera like *Roseburia* that some studies find to be increased in CRC but other studies report either decreased or no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera. Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of CRC microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue directly may provide clearer answers but is not without limitations. The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of CRC. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of CRC. We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of CRC. Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionaly accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.

### 2018-01-25 submission/manuscript.tex
ada background edits

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been purported to play an active role in disease
pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying biomarkers or individual microbes but did so on a small
total number of individuals and only investigated stool.

Although there has been an intense focus on microbiota-based biomarker
discovery for CRC, the candidate genera seem to be numerous. Some
studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
some of these studies mouth-associated genera are far from the only
microbes identified to be associated with CRC. These other genera
include, but are not necessarily limited to, \emph{Providencia},
\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
\emph{Escherichia/Shigella} and \emph{Streptococcus} also have good
\emph{in vivo} evidence that individual species within the genus can be
important in the pathogenesis of CRC {[}5,19,20{]}. Other studies have
also identified \emph{Akkermansia muciniphila} and \emph{Bacteroides
fragilis} as potential additional markers of CRC with \emph{Bacteroides
fragilis} having good mechanistic studies into how this may occur
{[}15,21,22{]}. Still, other studies have identified genera like
\emph{Roseburia} as increased in CRC while others have reported it to be
decreased or have found no difference in any genera at all
{[}15,18,23,24{]}.

Most of these studies have focused on carcinoma but the adenoma
observations are not any clearer at identifying candidate genera. Some
groups have focused on more broad scale community metrics and have found
that metrics such as richness are decreased in adenoma versus controls
while others have identified \emph{Lactococcus}, \emph{Pseudomonas},
\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, it only analyzed a total of 509 stool samples and with some
recent studies, that were not included, matching this sample size alone,
it is hard to know how extrapolatable these findings truly are. That
particular meta-analysis also added more potential microbial
associations to both carcinoma (\emph{Pantoea agglomerans}
\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
\emph{Lactobacillus}) that need to be investigated further, since a
number of these genera have been found to be enriched in controls and
not lesion (adenoma or carcinoma) {[}13,16,17{]}.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}18,28{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}29{]}. Due to these many differences that could arise
between stool and tissue, an additional question our meta-analysis aims
to answer is whether there are consistent patterns that emerge across
studies regardless of whether they used stool or tissue samples.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, re-examines important
genera across adenoma and carcinoma across study, and examines
differences and similarities between stool and tissue microbiota in the
context of CRC. Importantly, this analysis and approach could provide
valuable insights into the common genera that are both protective and
detrimental in CRC and whether broad bacterial community measurements
can account for these changes that were not previously provided by
earlier meta-analysis studies {[}11{]}.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been purported to play an active role in disease
pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
been numerous case/control studies investigating the microbiota in the
formation of both adenoma and carcinoma. A recent meta-analysis
investigated whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. This meta-analysis focused
on identifying microbial signatures of CRC (biomarkers) but did so on a
small total number of individuals and only investigated stool.

Although there has been an intense focus on microbiota-based biomarker
discovery for CRC, the number of candidate genera seem to be endless.
Some studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
these studies, mouth-associated genera are far from the only microbes
identified to be associated with CRC. These other genera include, but
are not limited to, \emph{Providencia}, \emph{Mogibacterium},
\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential markers of CRC with good
mechanistic studies for the latter {[}15,21,22{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, the sample size (n = 509) is equal to or less than some more
recent individual studies investigating the microbiota and CRC, making
it hard to know how extrapolatable these findings are. That particular
meta-analysis also added more potential microbial associations to both
carcinoma (\emph{Pantoea agglomerans} \emph{Ruminococcus},
\emph{Lactobacillus}) and adenoma (\emph{Prevotella},
\emph{Methanosphaera}, \emph{Succinovibrio}, \emph{Haemophilus
parainfluenzae}, \emph{Ruminococcus}, \emph{Lactobacillus}) stages of
CRC that need to be investigated further, since a number of these genera
have been found to be enriched in controls and not disease
{[}13,16,17{]}. There are also genera like \emph{Roseburia} that some
studies find to be increased in CRC but other studies report either
decreased or no difference {[}15,18,23,24{]}.

Most of these studies have focused on carcinoma but the adenoma
observations are not any clearer at identifying candidate genera. Groups
focusing on broad scale community metrics have found that metrics such
as richness are decreased in the adenoma stage of CRC versus controls.
Other studies have identified \emph{Lactococcus}, \emph{Pseudomonas},
\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. Based on
the studies mentioned, there seems to be very little overlap between the
genera identified to be associated with adenoma and carcinoma, with
\emph{Lactobacillus} being one of the few commonalities.

Targeting the identification of CRC microbial biomarkers within stool
seems logical since it offers an easy and cost-effective way to stratify
risk and the current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not without risk of complications. Although stool
represents an easy and less invasive way to assess risk, it is not clear
how well this sample reflects adenoma- and carcinoma- associated
microbial communities. Some studies have tried to assess this in health
and disease but are limited by their sample size {[}18,28{]}. Sampling
the microbiota directly associated with colon tissue directly may
provide clearer answers but is not without limitations. The community
present for sampling following the colonoscopy bowel prep may reflect
the better adhered microbiota versus the resident community.
Additionally, these samples contain more host DNA, potentially limiting
the types of analysis that can be done. It is well known that low
biomass samples can be very difficult to work with and results can be
study dependent due to the randomness of contamination {[}29{]}.

In comparison to the previous meta-analysis, this study significantly
increases the total stool samples investigated, re-examines important
genera across adenoma and carcinoma across study, and examines
differences and similarities between stool and tissue microbiota in the
context of CRC. Importantly, this analysis and approach could provide
valuable insights into the common genera that are both protective and
detrimental in CRC and whether broad bacterial community measurements
can account for these changes that were not provided by earlier
meta-analysis studies {[}11{]}.

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of CRC. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased
relative risk (RR) for adenoma or carcinoma stages of CRC. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma stages of CRC. Next, using Random Forest models, we analyzed
whether the full community or only the combined top 5 increased and top
5 decreased RR genera resulted in better model classification, based on
the area under the curve (AUC). Finally, we also examined at what effect
and sample size the studies used were powered for and the sample size
needed to get to the traditionaly accepted 80\% power. Our results from
these analyses suggests that the bacterial community changes as disease
severity worsens, that specific members are important for disease
classification, and that many of the studies are underpowered for
assessing small effect sizes.
